---
title: "04E - R basics 4"
---



# R Lists 

Lists will allow us to store a variety of data structures under a single variable. This means we could store a vecor,matrix, data frame, etc. under a single list. For example:


Create a vector
```{r}
v <- c(1,2,3,4,5)
```


Create a matrix
```{r}
m <- matrix(1:10,nrow=2)
```


Create a data frame
```{r}
df <- women
```


# Using list()

```{r}
li <- list(v,m,df)
li
```

You will notice that the list() assigned numbers to each of the objects in the list, but we can also assign names in the following manner:

```{r}
li <- list(sample_vec = v,sample_mat = m, sample_df = df)
li
```


#Selecting objects from a list

Single brackets
```{r}
li[1] # By index
```

```{r}
li['sample_vec'] # By name
```


```{r}
li[['sample_vec']]
```

```{r}
li$sample_vec # Can also use $ notation
```

You can also index on top of this once you've selected the object from the list, for example:

```{r}
li[['sample_vec']][1] # Second set of indexing
```

```{r}
li[['sample_mat']][1:2,1:2]
```

```{r}
li[['sample_df']]['height']
```
